### 2021-12-14 thesis/tex/abstract.tex
update

The failure of the Standard Model of particle physics to predict neutrino masses forces us to amend it. We have no reason to believe that 
the interactions currently described by the gauge group are the whole picture, nor should we expect that the number of fermions hitherto observed is correct.
In this thesis, we explore two modifications to the Standard Model, and examine whether these amendments are consistent with measured data.
Firstly, we consider the sterile neutrino, which have no interactions described by the Standard Model. We examine the effect of this 
new fermion on the neutrino oscillation probabilities, and present what would be a detectable signal in the \si{\TeV} range.
Secondly, we consider interactions beyond the Standard Model, possibly stemming from a higher-order theory.
We show how the parameters of these Non-Standard Interactions (NSI) can modify the oscillation probabilities, and within which energy range we expect to be able to discern this signal.
We use data from and simulate event counts in two Cherenkov detectors: IceCube and DeepCore. Moreover, we generate data and simulate a proposed upgrade of the DeepCore detector: PINGU. 
Using IceCube track events, we obtain best-fit values $\dm[41] = \SI{0.01}{\eV^2}$ and $\theta_{24} = 0.67$ for our sterile neutrino hypothesis at 
a p-value of $20\%$, which is not statistically significant. Hence, we found no evidence of a sterile neutrino in IceCube data. 
Moreover, were unable to distinguish a signal from $\theta_{34}$ in our IceCube simulation.
We obtain stringent bounds on the NSI parameters, and compare those to previous results. We show that PINGU is expected to further narrow the bound 
on $\emt$, especially the considering a joint analysis together with IceCube and DeepCore. Finally, we see that an anti-correlation between 
$\eem$ and $\eet$ at probability level was propagated down to event level, which we expect to be observable by PINGU.

>>>>

The failure of the Standard Model of particle physics to predict neutrino masses invites us to amend it. We have no reason to believe that 
the interactions currently described by the gauge group are the whole picture, nor should we expect that the number of fermions hitherto observed is correct.
In this thesis, we explore two amendments to the Standard Model, and examine whether either of these are consistent with measured data.

Firstly, we consider the sterile neutrino, which have no interactions described by the Standard Model. We examine the effect of this 
new fermion on the neutrino oscillation probabilities, and present what could be a detectable signal in the \si{\TeV} energy range.
Secondly, we consider interactions beyond the Standard Model, possibly stemming from a higher-order theory.
We show how the parameters of these Non-Standard Interactions (NSI) can modify the oscillation probabilities, and within which energy range we expect to be able to discern this signal.
We use data from and simulate event counts in two Cherenkov detectors: IceCube and DeepCore. Moreover, we generate data and simulate a proposed upgrade of the DeepCore detector: PINGU.

Using IceCube track events, we obtain best-fit values $\dm[41] = \SI{0.01}{\eV^2}$ and $\theta_{24} = 0.67$ for our sterile neutrino hypothesis at 
a p-value of $20\%$, which is not statistically significant. Hence, we found no evidence of a sterile neutrino in IceCube data. 
Moreover, were unable to distinguish a signal from $\theta_{34}$ in our IceCube simulation.
We obtain stringent bounds on the NSI parameters, and compare those to previous results in literature. We show that PINGU is expected to further narrow the bound 
on $\emt$, especially the considering a joint analysis together with IceCube and DeepCore. Finally, we see that an anti-correlation between 
$\eem$ and $\eet$ at probability level was propagated down to event level, which we expect to be observable by PINGU.

### 2021-12-14 thesis/tex/nsiResults.tex
update

\subsection{$\chi^2$ Minimization}\label{sec:nsi_chi2}
For our analyses, we define our $\chi^2$ as
\begin{align}\label{eq:chisq_NSI}
    \chi^{2}(\hat{\theta},\alpha,\beta, \kappa)=\sum_{ijk} \frac{\left(N^\text{th}-N^\text{data}\right)_{ijk}^{2}}
    {\left(\sigma^\text{data}_{ijk}\right)^{2} + \left(\sigma^\text{syst}_{ijk}\right)^{2}}+ 
    \frac{(1-\alpha)^2}{\sigma_\alpha^2} + \frac{\beta^2}{\sigma_\beta^2}\,
\end{align}
where we minimize over the model parameters $\hat{\theta} \in \{\dm, \theta_{23}, \epsilon\}$, the penalty terms $\alpha$ and $\beta$, and the free parameter $\kappa$.
$N_{ijk}^\text{th}$ is the expected number of events from theory in bin $\{i,j,k\}$, where $i$ denotes the $\Ereco$ bin, $j$ denotes the $\zreco$ bin,
and $k$ denotes the event-type bin, i.e. track or cascade. 
$N_{ijk}^\text{data}$ is the observed number of events in that bin. $\sigma^\text{data}_{ijk}$ is the experimental uncertainty, and $\sigma^\text{syst}_{ijk}$ the uncorrelated systematic
uncertainty.

In our simulations of $N_{ijk}^\text{th}$, we set all standard oscillation parameters to their current best-fit values of Eq.~\ref{eq:PINGUparams}, 
except for $\dm$ and $\theta_{23}$
which we vary over their $3\sigma$ limits \SIrange{2.435e-3}{2.598e-3}{\eV\squared} and \SIrange{40.1}{51.7}{\degree}, respectively.

@@

\subsection{Constraining the NSI parameters}\label{sec:constraining_NSI}
After the oscillation parameters have been marginalized out, we plot $\Delta \chi^2$ for each of the four NSI parameters in Fig.~\ref{fig:3D_NO}. 
The results are shown in Fig.~\ref{fig:IC_3D} and summarized in Tables~\ref{table:IC_DC_results} and~\ref{table:PINGU_joint_results}. Each region is bounded
by the best and worst case scenario, as defined in Table~\ref{table:syst_errors}, while the middle line is the baseline scenario. We again emphasize that 
no scenarios are considered for DeepCore, since they already provided in the data release~\cite{DC2019data}.

>>>>

\subsection{\texorpdfstring{$\chi^2$}{Chi-squared} Minimization}
\label{sec:nsi_chi2}
For our analyses, we define our $\chi^2$ as
\begin{align}\label{eq:chisq_NSI}
    \chi^{2}(\hat{\theta},\alpha,\beta, \kappa)=\sum_{ijk} \frac{\left(N^\text{th}-N^\text{data}\right)_{ijk}^{2}}
    {\left(\sigma^\text{data}_{ijk}\right)^{2} + \left(\sigma^\text{syst}_{ijk}\right)^{2}}+ 
    \frac{(1-\alpha)^2}{\sigma_\alpha^2} + \frac{\beta^2}{\sigma_\beta^2}\,
\end{align}
where we minimize over the model parameters $\hat{\theta} \in \{\dm, \theta_{23}, \epsilon\}$, the penalty terms $\alpha$ and $\beta$, and the free parameter $\kappa$.
$N_{ijk}^\text{th}$ is the expected number of events from theory in bin $\{i,j,k\}$, where $i$ denotes the $\Ereco$ bin, $j$ denotes the $\zreco$ bin,
and $k$ denotes the event-type bin, i.e. track or cascade. 
$N_{ijk}^\text{data}$ is the observed number of events in that bin. $\sigma^\text{data}_{ijk}$ is the experimental uncertainty, and $\sigma^\text{syst}_{ijk}$ the uncorrelated systematic
uncertainty.

In our simulations of $N_{ijk}^\text{th}$, we set all standard oscillation parameters to their current best-fit values of Eq.~\ref{eq:PINGUparams}, 
except for $\dm[31]$ and $\theta_{23}$
which we vary over their $3\sigma$ limits \SIrange{2.435e-3}{2.598e-3}{\eV\squared} and \SIrange{40.1}{51.7}{\degree}, respectively~\cite{nufit}.

@@

\subsection{Constraining the NSI parameters}\label{sec:constraining_NSI}
We let each of the four NSI parameters considered assume a value, while keeping the other three fixed at zero. We consider the following ranges of values for each parameter:
\begin{align}
   -0.07 \le &\, \ett \le 0.07 \nonumber \\
   -0.03 \le &\, \emt \le 0.03 \nonumber \\
   -0.3 \le &\, \eem \le 0.3 \nonumber \\
   -0.3 \le &\, \eet \le 0.3 \nonumber \\
\end{align}
After the oscillation parameters $\dm[31]$ and $\theta_{23}$ have been marginalized out, we plot $\Delta \chi^2$ for each of the four NSI parameters in Fig.~\ref{fig:3D_NO}. 
The results are shown in Fig.~\ref{fig:IC_3D} and summarized in Tables~\ref{table:IC_DC_results} and~\ref{table:PINGU_joint_results}. Each region is bounded
by the best and worst case scenario, as defined in Table~\ref{table:syst_errors}, while the middle line is the baseline scenario. We again emphasize that 
no uncertainty scenarios are considered for DeepCore, since they already provided in the data release~\cite{DC2019data}.

### 2021-12-14 thesis/tex/sterileResults.tex
update

\section{The Sterile Hypothesis}\label{sec:sterileResults}
With the null ($3\nu$) hypothesis normalized to the IceCube Monte Carlo as in Eq.~\ref{eq:MC_norm}, we are now in good shape to study the sterile neutrino effect on the probabilities, and how that compares to data
\footnote{We do not consider DeepCore nor PINGU in the sterile neutrino analysis because the $\Pamam$ resonance does not appear in the \si{\GeV} region for \si{\eV^2}-scale $\dm[41]$, as shown in Fig.~\ref{fig:sterile_resonance}.
We bring DeepCore and PINGU back in the next section.}.
First, let us see how the collected data from Fig.~\ref{fig:IC_data} deviates from the predicted $3\nu$ oscillations.
\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{figures/IC_rates.pdf}
    \caption{How the IceCube null hypothesis from~\cite{IC2020} compares to their collected data.}\label{fig:IC_rates}
\end{figure}
Fig.~\ref{fig:IC_rates} shows an impressive agreement to the standard $3\nu$ oscillation picture, with the largest deviations of $~6\%$
for neutrinos with high single-digit \si{\TeV} values. In zenith, the data only deviates $\pm 2\%$.

@@

By tuning the scaling $f$ of the uncorrelated systematic error, we can shift the contour. We extract the contour from the 
IceCube sterile neutrino analysis~\cite{IC2020}, and compare in Fig.~\ref{fig:error_tuning}. We see that the contour in the 
higher $\dm[41]$ region above \SI{1}{\eV\squared} is very sensitive to the tuning, while the lower region between \SIrange[]{1e-1}{1}{\eV\squared}
remains fixed. In both cases, we obtain best-fit values of $\dm[41] = \SI{0.01}{\eV^2}$ and $\theta_{24} = 0.67$ ($\sin^22\theta_{24} = 0.95$) at 
a p-value of $33\%$.
\begin{figure}
    \centering
    \includegraphics[scale=0.58]{figures/s24_error_tuning.pdf}
    \caption{The 90\% CL contour with two different systematic error scalings, compared 
    with the official IceCube result from~\cite{IC2020}.}\label{fig:error_tuning}
\end{figure}

@@

In Fig.~\ref{fig:final_rate_plot}, we see that the simulations are consistent with the 3+0 hypothesis with deviations of 6\% at high single-digit 
\si{\TeV} energies.
Wanting to investigate the effect of $\theta_{34}$, we also simulated the 3+1 hypothesis using $\theta_{24} = \theta_{34}$. However, we did not see 
a difference in the exclusion regions compared to the regions when $\theta_{34} = 0$. Thus, we conclude that on our work, IceCube was not able to 
distinguish any effect due to the $\theta_{34}$ parameter.

>>>>

\section{The Sterile Neutrino Hypothesis}\label{sec:sterileResults}
With the null ($3\nu$) hypothesis normalized to the IceCube Monte Carlo as in Eq.~\ref{eq:MC_norm}, we are now in good shape to study the sterile neutrino effect on the probabilities, and how that compares to data
\footnote{We do not consider DeepCore nor PINGU in the sterile neutrino analysis because the $\Pamam$ resonance does not appear in the \si{\GeV} region for \si{\eV^2}-scale $\dm[41]$, as shown in Fig.~\ref{fig:sterile_resonance}.
We bring DeepCore and PINGU back in the next section.}.
First, let us see how the collected data from Fig.~\ref{fig:IC_data} deviates from the predicted $3\nu$ oscillations.
\begin{figure}
    \centering
    \includegraphics[width=1\textwidth]{figures/IC_rates.pdf}
    \caption{How the IceCube null hypothesis from~\cite{IC2020} compares to their collected data.}\label{fig:IC_rates}
\end{figure}
Fig.~\ref{fig:IC_rates} shows an impressive agreement to the standard $3\nu$ oscillation picture, with the largest deviations of $~6\%$
for neutrinos at $\SI{10}{\TeV}$ values. In zenith, the data only deviates $\pm 2\%$.

@@

By tuning the scaling $f$ of the uncorrelated systematic error, we can shift the contour. We extract the contour from the 
IceCube sterile neutrino analysis~\cite{IC2020}, and compare in Fig.~\ref{fig:error_tuning}. We see that the contour in the 
higher $\dm[41]$ region above \SI{1}{\eV\squared} is very sensitive to the tuning, while the lower region between \SIrange[]{1e-1}{1}{\eV\squared}
remains fixed. Using the case of 15\% uncorrelated systematic error, we obtain best-fit values of $\dm[41] = \SI{0.01}{\eV^2}$ and $\theta_{24} = 0.67$ ($\sin^22\theta_{24} = 0.95$) at 
a p-value of $20\%$. This is not statistically significant, 
hence we find no evidence for a sterile neutrino within the parameter range $0.01 \le \dm[41] \le \SI{1}{\eV^2}, 0.01\le \sin(2\theta_{24})^2 \le 1$  in IceCube data.
\begin{figure}
    \centering
    \includegraphics[scale=0.58]{figures/s24_error_tuning.pdf}
    \caption{The 90\% CL contour with two different systematic error scalings, compared 
    with the official IceCube result from~\cite{IC2020}.}\label{fig:error_tuning}
\end{figure}

@@

Wanting to investigate the effect of $\theta_{34}$, we also simulated the 3+1 hypothesis using $\theta_{24} = \theta_{34}$. However, we did not see 
a difference in the exclusion regions compared to the regions when $\theta_{34} = 0$. Thus, we conclude that on our work, IceCube was not able to 
distinguish any effect due to the $\theta_{34}$ parameter. However, since our dataset only included muon track events in the \SIrange{0.5}{10}{\TeV} range,
one might be able to expect a larger effect of $\theta_{34}$ if one includes cascade events as well.
